---
title: "Economic Freedom Index 2019 — Descriptive Report"
author: "Grace Bianchi"
date: "2025-10-14"
output: pdf_document
---

# Introduction to the Economic Freedom Index 2019 data set

**Context**

The Economic Freedom Index is poised to help readers track over two decades of 
the advancement in economic freedom, prosperity, and opportunity and promote 
these ideas in their homes, schools, and communities. The Index covers 12 
freedoms – from property rights to financial freedom – in 186 countries. The 
central variable is 2019 Score, which summarizes the degree of economic freedom
with a score from 0-100. A higher score indicates greater economic freedom.

**Content**

The data (last updated 26/02/2019) is presented in CSV format as follows: 
CountryID, Country Name, WEBNAME, Region, World Rank, Region Rank, 2019 Score, 
Property Rights, Judical Effectiveness, Government Integrity, Tax Burden, Gov't 
Spending, Fiscal Health, Business Freedom, Labor Freedom, Monetary Freedom, 
Trade Freedom, Investment Freedom, Financial Freedom, Tariff Rate (%), Income 
Tax Rate (%), Corporate Tax Rate (%), Tax Burden % of GDP, Gov't Expenditure % 
of GDP , Country, Population (Millions), "GDP (Billions, PPP)", GDP Growth Rate
(%), 5 Year GDP Growth Rate (%), GDP per Capita (PPP), Unemployment (%), 
Inflation (%), FDI Inflow (Millions), Public Debt (% of GDP).

**Goal**

The goal of this analysis is to conduct a descriptive study of the Economic 
Freedom Index 2019 dataset to examine the factors that influence economic 
freedom across different nations and regions. By exploring the distribution of 
the 2019 scores and their associations with economic, institutional, and social 
variables, we aim to understand how economic freedom varies globally. The 
findings from this analysis can help inform policy decisions that promote 
higher levels of economic freedom and sustainable economic development 
worldwide.

**Research Questions**

The analysis will address the following key inquiries:

- What are the top 5 countries with the highest levels of economic freedom in 
2019?

- What are the 5 countries with the lowest levels of economic freedom in 2019?

- How does GDP per capita relate to economic freedom?

- Do countries with stronger property rights tend to exhibit higher economic 
freedom?

- What is the relationship between government integrity and economic freedom?

- Do countries with greater trade openness experience higher economic freedom?

- Are there regional differences in economic freedom scores across the world?

- What patterns or similarities emerge between countries with high and low 
economic freedom?

This analysis will provide an in-depth exploration of these questions, aiming to 
uncover patterns that can help guide policies to strengthen the institutional 
and economic conditions that foster greater economic freedom globally.

# Variable Description

The Economic Freedom Index 2019 dataset contains several key variables that aim
to quantify economic freedom and its contributing factors across various 
countries. Below is a description of the primary variables used in this 
analysis:

- Country Name: The name of each country included in the Economic Freedom Index.

- Region: The geographical region where the country is located.

- 2019 Score: The overall measure of economic freedom on a scale from 0 to 100,
where higher values indicate greater freedom.

- Property Rights: Index measuring the degree to which private property is
protected by the law and enforcement is effective.

- Judicial Effectiveness: Reflects the efficiency and fairness of the judical 
system, including independence of courts.

- Government Integrity: Indicates the level of transparency and freedom from 
corruption in government institutions.

- Trade Freedom: Measures the openness of a country's trade policy, based on 
tariffs and non-tariff barriers.

- GDP per Capita (PPP): Gros Domestic Product per person, adjusted for 
purchasing power parity, in international dollars.

- GDP Growth Rate (%): The percentage change in a country's GDP from the 
previous year.

- Population (Millions): Total population of the country, expressed in millions.

This variable description provides a detailed understanding of the dataset, 
allowing for a clearer interpretation of the analysis results.

# Package Installation and Loading

```{r}
# Install packages only if not already installed
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("dplyr")) install.packages("dplyr")
if (!require("ggpubr")) install.packages("ggpubr")
if (!require("reshape2")) install.packages("reshape2")
if (!require("janitor")) install.packages("janitor")

# Load libraries
library(tidyverse)
library(ggplot2)
library(dplyr)
library(ggpubr)
library(reshape2)
library(janitor)
library(corrplot)
```

```{r}
# Read the Economic Freedom Index Data

economic_freedom_data = read.csv("economic_freedom_index2019_data.csv")
```

# Descriptive Analysis

```{r, warning = FALSE}
# Changes data type from character to numeric

economic_freedom_data$X2019.Score <- as.numeric(economic_freedom_data$X2019.Score)
economic_freedom_data$Property.Rights <- as.numeric(economic_freedom_data$Property.Rights)
economic_freedom_data$Judical.Effectiveness <- as.numeric(economic_freedom_data$Judical.Effectiveness)
economic_freedom_data$Government.Integrity <- as.numeric(economic_freedom_data$Government.Integrity)
economic_freedom_data$Trade.Freedom <- as.numeric(economic_freedom_data$Trade.Freedom)
economic_freedom_data$GDP.per.Capita..PPP. <- as.numeric(gsub("[$,]", "", economic_freedom_data$GDP.per.Capita..PPP.))
economic_freedom_data$GDP.Growth.Rate.... <- as.numeric(economic_freedom_data$GDP.Growth.Rate....)
economic_freedom_data$Population..Millions. <- as.numeric(economic_freedom_data$Population..Millions.)
```

## Summary Statistics

```{r}
# Provides summary statistics for variables

summary_vars <- economic_freedom_data %>%
  select(X2019.Score, Property.Rights, Judical.Effectiveness, Government.Integrity,
         Trade.Freedom, GDP.per.Capita..PPP., GDP.Growth.Rate...., Population..Millions.)

summary(summary_vars)
```

## Distribution of Economic Freedom Index Scores
```{r, warning = FALSE}
ggplot(economic_freedom_data, aes(x = X2019.Score)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "black") +
  labs(
    title = "Distribution of Economic Freedom Scores (2019)",
    x = "Economic Freedom Score (0–100)",
    y = "Number of Countries"
  ) 

```

## Exploring Different Binwidths for the Histogram of Economic Freedom Scores

```{r, warning = FALSE}
ggplot(economic_freedom_data, aes(x = X2019.Score)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "black") +
  labs(
    title = "Distribution of Economic Freedom Scores (2019) (Binwdith = 0.5)",
    x = "Economic Freedom Score (0–100)",
    y = "Number of Countries"
  ) 

ggplot(economic_freedom_data, aes(x = X2019.Score)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(
    title = "Distribution of Economic Freedom Scores (2019) (Binwdith = 1)",
    x = "Economic Freedom Score (0–100)",
    y = "Number of Countries"
  ) 

ggplot(economic_freedom_data, aes(x = X2019.Score)) +
  geom_histogram(binwidth = 1.5, fill = "steelblue", color = "black") +
  labs(
    title = "Distribution of Economic Freedom Scores (2019) (Binwdith = 1.5)",
    x = "Economic Freedom Score (0–100)",
    y = "Number of Countries"
  )
```

## Density Plot of Economic Freedom Scores
```{r, warning = FALSE}
ggplot(economic_freedom_data, aes(x = X2019.Score)) +
  geom_density(fill = "lightblue") +
  labs(
    title = "Density Plot of Economic Freedom Scores (2019)",
    x = "Economic Freedom Score (0–100)",
    y = "Density"
  ) 
```

**Conclusions**

**Histogram Analysis with Different Binwidths:**

Binwidth = 0.5: The historgram shows a more detailed view which reveals small
fluctuations in the distribution of economic freedom scores. It would be most 
useful when exploring fine-grained patterns or checking for small clusters or 
gaps in the data. Binwidth = 1: The histogram provides a clearer view of the 
overall shape of the distribution without excessive noise. It would be most 
useful when exploring descriptive summaries and initial visual analyses. 
Binwidth = 1.5: The histogram becomes smoother, making overall trends, such as
the peak around the gobal mean and the slight right skew, easier to see. Most 
countries scores exist from 50 to 75 with less countries scoring outside that 
range. It would be most useful when presenting general patterns or summarizing 
findings.

**Density Plot Analysis:**

The density plot provides a smooth estimate of the distrbution of economic 
freedom scores. It shows that the bulk of countries are centered around a score
of approximately 60, with the distribution gradually tapering off toward both 
higher and lower values. The plot confirms that the distribution is slightly
right-skewed, indicating that there are relatively few countries with very high 
freedom scores (80+), while a large number of countries fall within the moderate
range (50-70).

**Insights and Implications:**

**Economic Freedom Concentration:** The histograms and density plot reveal that 
the majority of countries have economic freedom scores clustered between 50 and 
70, suggesting that the global economic freedom is generally moderate. 
Relatively few countries fall at the extremes of the distribution, with only a 
handful achieving very high scores above 80 or very low scores below 40.

**Right-Skewed Distribution:** The distribution is slightly right-skewed, 
indicating that while some nations experience exceptionally high levels of
economic freedom, many more remain below those top levels. This pattern may 
reflect structural or institutional challenges, such as weak property rights, 
government intervention, or limited judicial effectiveness. These challenges
may prevent countries from reaching the highest tiers of freedom.

**Policy Implications:**

Since most countries cluster in the mid-range of economic freedom, policy 
efforts should prioritize institutional and regulatory reforms that enhance
transparency, strengthen judicial independence, and promote open trade.
Incremental improvements in these areas could help shift the bulk of nations
toward higher freedom scores and, in turn, stimulate greater economic growth
and opportunity. 

# GDP Per Capita by Region

```{r, warning = F}
# Create logged GDP per capita variable
economic_freedom_data <- economic_freedom_data %>%
  mutate(Logged_GDP_per_Capita = log(GDP.per.Capita..PPP.))

# Boxplot of GDP per Capita by Region 
ggplot(economic_freedom_data,
       aes(x = reorder(Region, -Logged_GDP_per_Capita),
           y = Logged_GDP_per_Capita, fill = Region)) +
  geom_boxplot() +
  labs(
    title = "GDP per Capita by Region",
    x = "Region",
    y = "Logged GDP per Capita"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

From the boxplot analysis of GDP per capita by region, we can draw the following conclusions:

**Significant Regional Difference:** There are clear differences in GDP per 
capita across regions. Europe and the Middle East and North Africa have the 
highest median GDP per capita, indicating that wealthier economies are 
concentrated in these regions. Sub-Saharan Africa, by contrast, has the lowest
median, suggesting limited economic development relative to other parts of the
world.

**High Variability in Some Regions:** The Asia-Pacific and Middle East and 
North Africa regions display the widest spreads, showing that while some 
countries (e.g., Singapore, Qatar) are very wealthy, others in the same region
remain far below the regional averages. This reflects substantial within-region
economic inequality.

**Regions with Lower GDP per Capita:** Sub-Saharan Africa stands out with a 
consistently lower GDP per capita and a wide range, implying both lower overall
wealth and uneven development among its countries. The Americas show mid-range 
values, with both high-income and developing economic represented.

**Outliers:** A few outliers appear above the main boxes, representing countries
with exceptionally high GDP per capita relative to their regional peers (likely
advanced financial or resource-driven economies).

**Conclusion:** The boxplot highlights pronounced global inequality in income 
distribution across regions. Wealthier regions such as Europe and the Middle 
East cluster toward higher GDP per capita, while Sub-Saharan Africa remains 
substantially below. The broad spread in regions like Asia-Pacific suggests 
diverse levels of deelopement, combining both emerging and advanced economies. 
These findings provide important context for interpreting regional differences 
in economic freedom, as wealth and institutional strength often reinforce each 
other.

# Correlation Analysis

```{r, warning = F}
cor_matrix <- economic_freedom_data %>%
  select(X2019.Score,
         Property.Rights,
         Judical.Effectiveness,
         Government.Integrity,
         Trade.Freedom,
         GDP.per.Capita..PPP.,
         GDP.Growth.Rate....,
         Population..Millions.) %>%
  cor(use = "complete.obs")

# Round to 2 decimals
round(cor_matrix, 2)
```

**Correlation Analysis with Corrplot**

```{r, warning = F}
corrplot(cor_matrix,
         method = "circle",
         type = "upper",
         tl.col = "black",
         tl.srt = 45,
         title = "Correlation Matrix of Economic Freedom Factors (2019)",
         mar = c(0, 0, 1, 0))
```

**Correlation Analysis and Conclusions**

**Interpretation of the Correlation Analysis:**

Property Rights and Economic Freedom:

There is a strong positive correlation (r = 0.86) between Property Rights and
Economic Freedom. This suggests that nations with well-established legal
protections for private ownership, contact enforcement, and secure property
sysems tend to experience higher levels of overall economic freedom. The 
protection of property rights appears to be one of the strongest institutional 
pillars supporting free markets.

Judicial Effectiveness and Economic Freedom: 

A high positive correlation (r = 0.81) indicates that countries with efficient 
and impartial judicial systems tend to have greater economic freedom. Judicial 
effectiveness reinforces the rule of law by ensuring contracts are honored and 
corruption is limited, fostering investor confidence and fair competition.

Government Integrity and Economic Freedom:

The correlation between Government Integrity and Economic Freedom is strong 
(r = 0.80). This demonstrates that transparency and low corruption levels are 
closely tied to freer economic systems. When citizens and businesses can trust
public institutions, economic activity is less distorted by rent-seeking and 
political favoritism.

Trade Freedom and Economic Freedom:

The relationship between Trade Freedom and overall Economic Freedom is 
moderately strong (r = 0.71). Countries that maintain open trade policies, such
as lower tariffs, minimal quotas, and fewer regulatory barriers, tend to perform
better in terms of overall economic freedom. This reinforces the idea that 
integration into global markets supports freer domestic economies.

GDP per Capita (PPP) and Economic Freedom:

There is a moderately positive correlation (r = 0.64) between GDP per capita and
Economic Freedom. Wealthier nations generally enjot higher economic freedom, 
reflecting how sustained prosperity and robust institutions reinforce one 
another. However, the relationship is not perfect. Some resource-rich or 
reforming nations may have higher GDP levels without matching levels of 
institutional freedom.

GDP Growth Rate and Economic Freedom:

A weak positive correlation (r = 0.18) suggests that short-term economic growth 
does not strongly predict higher economic freedom. While freer economies may 
grow more steadily in the long term, annual growth rates can fluctuate due to 
external shocks, reforms, or commodity price cycles.

Population and Economic FreedomL

There is virutally no correlation (r = -0.04) between population size and 
economic freedom. This implies that demographic size alone does not determine 
how free an economy is. Both small and large nations can achieve either strong 
or weak institutional performance depending on their governance structures.

**Conclusions from the Correlation Analysis**

The results indicate that institutional quality variables, particularly Property 
Rights, Judicial Effectiveness, and Government Integrity, are the strongest 
correlates of economic freedom. Trade Freedom and GDP per Capital also show 
meaningful positive relationships, suggesting that open markets and wealth 
accumulation reinforce on another. By contrast, GDP Growth Rate and Population 
exhibit weak or negligible correlations, highlighting that economic freedom 
depends more on long-term institutional foundations than on short-term economic performance or demographic factors.

**Policy Implications**

These results highlight the importance of strengthening institutions to enhance
economic freedom. Policies that protect property rights, improve judicial 
effectiveness, reduce corruption, and promote open trade are most effective in
fostering long-term prosperity. Ultimately, strong and transparent governance
services as the foundation for sustainable economic growth and greater overall
freedom.

## Distribution of Property Rights Scores
```{r, warning = FALSE}
ggplot(economic_freedom_data, aes(x = Property.Rights)) +
  geom_histogram(binwidth = 2, fill = "plum", color = "black") +
  labs(
    title = "Distribution of Property Rights Scores",
    x = "Property Rights Score (0–100)",
    y = "Number of Countries"
  ) 

```

## Exploring Different Binwidths for the Histogram of Property Rights Scores

```{r, warning = FALSE}
ggplot(economic_freedom_data, aes(x = Property.Rights)) +
  geom_histogram(binwidth = 1, fill = "plum", color = "black") +
  labs(
    title = "Distribution of Property Rights Scores (binwidth = 1)",
    x = "Property Rights Score (0–100)",
    y = "Number of Countries"
  ) 

ggplot(economic_freedom_data, aes(x = Property.Rights)) +
  geom_histogram(binwidth = 1.75, fill = "plum", color = "black") +
  labs(
    title = "Distribution of Property Rights Scores (binwidth = 1.75)",
    x = "Property Rights Score (0–100)",
    y = "Number of Countries"
  ) 

ggplot(economic_freedom_data, aes(x = Property.Rights)) +
  geom_histogram(binwidth = 2.25, fill = "plum", color = "black") +
  labs(
    title = "Distribution of Property Rights Scores (binwidth = 2.25)",
    x = "Property Rights Score (0–100)",
    y = "Number of Countries"
  ) 
```

\newpage

## Density Plot of Property Rights Scores
```{r, warning = FALSE}
ggplot(economic_freedom_data, aes(x = Property.Rights)) +
  geom_density(fill = "mediumorchid") +
  labs(
    title = "Density Plot of Property Rights Scores",
    x = "Property Rights Score (0–100)",
    y = "Density"
  ) 
```

**Conclusions**

**Histogram Analysis with Different Binwidths:**

Binwidth = 1: The histogram shows a more detailed view of the Property Rights 
distributions, showing small fluctuations and minor peaks across the score 
range. However, the high granularity makes it harder to see the overall shape of 
the distribution. Binwidth = 1.75: The histogram strikes a balance between 
precision and clarity. It highlights that most countries cluster between 40 and 
65 on the Property Rights scale, with fewer countries at either extreme. 
Binwidth = 2.25: The histogram smooths out variation and reveals the main 
pattern that Property Rights scores are concentrated around mid-range values. 
Only a small number of countries achieve a very high (above 80) or very low 
(below 25) score.

**Density Plot Analysis:**

The density plot provides a smooth representation of the Property Rights 
distribution. It shows a clear concentration of countries around the mid-range 
(40-60), tapering off toward both ends. The slight double-peaked shape suggests 
a possible bimodal pattern, indicating the while some nations have moderately 
strong property protections, others remain considerably weaker in safeguarding 
ownership and enforcement rights.

**Insights and Implications:**

**Property Rights Concentration:** The histograms and density plot reveal that 
the majority of countries maintain moderate property rights with scores 
clustered between 40 and 60. This suggests that while property ownership is 
broadly recognized, consistent enforcement and legal protection still lag 
globally.

**Outliers:** A smaller number of nations exhibit very low or high scores, 
reflecting stark institutional contrasts. Countries range from weak legal 
systems and corruption issues to highly developed judicial protections.

**Policy Implications:** Strengthening property rights should be a key 
institutional priority especially in developing economies. Policies aimed at 
improving judicial enforcement, transparency, and corruption control can help 
nations move toward stronger, more reliable property systems that foster 
improvement and economic freedom.

**Explanation of the Two Modes:**

The presence of two modes (or peaks) in the density plot of property right 
scores suggests that the distribution of property right scores is bimodal. This 
indicates that there are two distinct groups of countries with different typical 
property right score levels. Here’s how to interpret this:

First Mode (Lower Property Right Scores): The first peak, typically around 
scores of 30-45, represents countries with weaker property right institutions. 
These nations often struggle with inconsistent legal enforcement, high levels 
of corruption, or unclear land ownership systems. In such environment, citizens
and businesses may face difficulties securing ownership titles or resolving
disputes, discouraging both domestic and foregin investment. Contributing 
factors may include: weak judicial systems and limited rule of law, government 
interference or expropriation risks, inefficient land registries or bureaucratic
delays.

Second Mode (Higher Property Right Scores): The second peak, typically appearing 
between 65-80, represents countries with well-established and consistently 
enforce property rights. These nations tend to have transparent legal systems, 
efficient courts, and stable governments that uphold ownership and contractual 
agreements. Commmon characteristics include: strong judicial independence and 
anti-corruption frameworks, secure land tenure and low expropriation risk, 
institutional reliability that supports long-term economic growth.

# Research Questions

**What are the top 5 countries with the highest levels of economic freedom in**
**2019?**

```{r, warning = F}
# Selecting the top 5 countries

top5 <- economic_freedom_data %>%
  arrange(desc(X2019.Score)) %>%
  select(Country.Name, X2019.Score) %>%
  head(5)

top5
```

```{r, warning = F}
# Visualizing the top 5 countries

ggplot(top5, aes(x = reorder(Country.Name, X2019.Score), y = X2019.Score, fill = Country.Name)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = X2019.Score), vjust = -0.5, color = "black", size = 4) +
  theme_minimal() +
  labs(
    title = "Top 5 Countries by Economic Freedom (2019)",
    x = "Country",
    y = "Economic Freedom Score (0–100)"
  )

```

**What are the top 5 countries with the lowest levels of economic freedom in**
**2019?**

```{r, warning = F}
# Selecting the bottom 5 countries

bottom5 <- economic_freedom_data %>%
  arrange(X2019.Score) %>%
  select(Country.Name, X2019.Score) %>%
  head(5)

bottom5
```

```{r, warning = F}
# Visualizing the bottom 5 countries

ggplot(bottom5, aes(x = reorder(Country.Name, X2019.Score), y = X2019.Score, fill = Country.Name)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = X2019.Score), vjust = -0.5, color = "black", size = 4) +
  theme_minimal() +
  labs(
    title = "Top 5 Countries with Lowest Economic Freedom (2019) Score",
    x = "Country",
    y = "Economic Freedom Score (0–100)"
  )
```

**How does GDP per capita relate to economic freedom?**

```{r, warning = F}
# Finding correlation between GDP per capita and economic freedom

gdp_economic_freedom_cor = cor(economic_freedom_data$X2019.Score,
    economic_freedom_data$GDP.per.Capita..PPP.,
    use = "complete.obs")

gdp_economic_freedom_cor
```

```{r, warning = F}
# Plotting scatterplot of GDP per capita vs economic freedom

gdp_freedom_plot <- ggplot(economic_freedom_data,
                           aes(x = GDP.per.Capita..PPP., y = X2019.Score)) +
  geom_point(aes(color = Region), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "GDP per Capita (PPP) vs Economic Freedom Score (2019)",
    x = "GDP per Capita (PPP, in $)",
    y = "Economic Freedom Score (0–100)",
    caption = "Source: Heritage Foundation, Economic Freedom Index 2019"
  ) +
  theme_minimal()

gdp_freedom_plot
```

**Statistical Summary of the Relationship**

We can further analyze the relationship by dividing the countries into groups 
based on their GDP per capita. This will allow us to compare happiness scores 
across wealth tiers.

```{r, warning = F}
# Create GDP per capita categories (quartiles)
economic_freedom_data <- economic_freedom_data %>%
  mutate(GDP_Category = cut(GDP.per.Capita..PPP.,
                            breaks = quantile(GDP.per.Capita..PPP.,
                                              probs = seq(0, 1, 0.25),
                                              na.rm = TRUE),
                            include.lowest = TRUE,
                            labels = c("Low", "Lower-Middle",
                                       "Upper-Middle", "High")))

# Summary of Economic Freedom by GDP category
gdp_freedom_summary <- economic_freedom_data %>%
  group_by(GDP_Category) %>%
  summarise(
    Mean_Freedom = mean(X2019.Score, na.rm = TRUE),
    SD_Freedom = sd(X2019.Score, na.rm = TRUE),
    Min_Freedom = min(X2019.Score, na.rm = TRUE),
    Max_Freedom = max(X2019.Score, na.rm = TRUE)
  )

# Display the summary table
gdp_freedom_summary
```

**Visualization by GDP Category**

To further understand how GDP per capita influences happiness, let’s visualize 
the happiness scores across different GDP categories.

```{r, warning = F}
ggplot(economic_freedom_data,
       aes(x = GDP_Category, y = X2019.Score, fill = GDP_Category)) +
  geom_boxplot() +
  labs(
    title = "Economic Freedom Score by GDP Category",
    x = "GDP Category",
    y = "Economic Freedom Score (0–100)",
    fill = "GDP Category"
  ) +
  theme_minimal()
```

**Conclusion**

The results of the analysis show a moderately strong positive correlation
(r = 0.637) between GDP per capita and economic freedom. This indicates that 
countries with higher income levels generally tend to exhibit greater economic 
freedom. The scatterplot clearly illustrates this upward trend, although some 
variability remains—suggesting that while economic prosperity and freedom often 
go hand in hand, GDP alone does not fully determine a nation’s level of economic
freedom.

The summary statistics and boxplot further reinforce this relationship: nations
in the High GDP category have the highest average economic freedom scores, while
those in the Low GDP group score noticeably lower. However, the presence of 
variation within each GDP tier implies that additional institutional and policy 
factors, such as governance integrity, trade openness, and property rights, also 
play key roles in shaping economic freedom.

**Do countries with stronger property rights tend to exhibit higher economic** 
**freedom?**

```{r, warning = F}
# Calculate the correlation between Property Rights and Economic Freedom
property_rights_freedom_correlation = cor(economic_freedom_data$Property.Rights,
                                          economic_freedom_data$X2019.Score,
                                          use = "complete.obs")

# Display the correlation value
property_rights_freedom_correlation
```

```{r, warning = F}
# Scatter plot: Property Rights vs Economic Freedom
ggplot(economic_freedom_data, aes(x = Property.Rights, y = X2019.Score)) +
  geom_point(color = "mediumorchid", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Property Rights vs Economic Freedom Score (2019)",
    x = "Property Rights Score (0–100)",
    y = "Economic Freedom Score (0–100)",
    caption = "Source: Heritage Foundation, Economic Freedom Index 2019"
  ) +
  theme_minimal()
```

```{r, warning = F}
# Create categories based on Property Rights scores (split into quintiles)
economic_freedom_data = economic_freedom_data %>%
  mutate(Property_Rights_Category = cut(Property.Rights,
                                        breaks = quantile(Property.Rights, probs = seq(0, 1, 0.2), na.rm = TRUE),
                                        include.lowest = TRUE,
                                        labels = c("Low Property Rights",
                                                   "Lower-Middle Property Rights",
                                                   "Upper-Middle Property Rights",
                                                   "High Property Rights",
                                                   "Very High Property Rights")))

# Summary of economic freedom scores by property rights category
property_rights_freedom_summary = economic_freedom_data %>%
  group_by(Property_Rights_Category) %>%
  summarise(
    Mean_Freedom = mean(X2019.Score, na.rm = TRUE),
    SD_Freedom = sd(X2019.Score, na.rm = TRUE),
    Min_Freedom = min(X2019.Score, na.rm = TRUE),
    Max_Freedom = max(X2019.Score, na.rm = TRUE)
  )

# Display the summary
property_rights_freedom_summary

```

```{r, warning = F}
# Boxplot of Economic Freedom by Property Rights Category
ggplot(economic_freedom_data, aes(x = Property_Rights_Category, y = X2019.Score, fill = Property_Rights_Category)) +
  geom_boxplot() +
  labs(
    title = "Economic Freedom Score by Property Rights Category (2019)",
    x = "Property Rights Category",
    y = "Economic Freedom Score (0–100)",
    fill = "Property Rights Category"
  ) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

**What is the relationship between government integrity and economic freedom?**

```{r, warning = F}
# Calculate the correlation between Government Integrity and Economic Freedom
gov_integrity_freedom_correlation = cor(economic_freedom_data$Government.Integrity,
                                        economic_freedom_data$X2019.Score,
                                        use = "complete.obs")

# Display the correlation value
gov_integrity_freedom_correlation
```

```{r, warning = F}
# Scatter plot: Government Integrity vs Economic Freedom
ggplot(economic_freedom_data, aes(x = Government.Integrity, y = X2019.Score)) +
  geom_point(color = "darkgreen", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Government Integrity vs Economic Freedom Score (2019)",
    x = "Government Integrity Score (0–100)",
    y = "Economic Freedom Score (0–100)",
    caption = "Source: Heritage Foundation, Economic Freedom Index 2019"
  ) +
  theme_minimal()
```

```{r, warning = F}
# Create categories based on Government Integrity (quintiles)
economic_freedom_data <- economic_freedom_data %>%
  mutate(GovIntegrity_Category = cut(Government.Integrity,
                                     breaks = quantile(Government.Integrity, probs = seq(0, 1, 0.2), na.rm = TRUE),
                                     include.lowest = TRUE,
                                     labels = c("Low Government Integrity",
                                                "Lower-Middle Government Integrity",
                                                "Upper-Middle Government Integrity",
                                                "High Government Integrity",
                                                "Very High Government Integrity")))

```

```{r, warning = F}
# Summary of Economic Freedom by Government Integrity category
gov_integrity_freedom_summary <- economic_freedom_data %>%
  group_by(GovIntegrity_Category) %>%
  summarise(
    Mean_Freedom = mean(X2019.Score, na.rm = TRUE),
    SD_Freedom   = sd(X2019.Score, na.rm = TRUE),
    Min_Freedom  = min(X2019.Score, na.rm = TRUE),
    Max_Freedom  = max(X2019.Score, na.rm = TRUE)
  )

# Display the summary
gov_integrity_freedom_summary

```

```{r, warning = F}
# Boxplot of Economic Freedom by Government Integrity category
ggplot(economic_freedom_data,
       aes(x = GovIntegrity_Category, y = X2019.Score, fill = GovIntegrity_Category)) +
  geom_boxplot() +
  labs(
    title = "Economic Freedom Score by Government Integrity Category (2019)",
    x = "Government Integrity Category",
    y = "Economic Freedom Score (0–100)",
    fill = "Government Integrity Category"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

```

Conclusion

The results of the analysis show a strong positive correlation between 
government integrity and economic freedom. The scatterplot indicates that 
countries with higher levels of government integrity tend to have higher 
economic freedom scores, reflecting the role of transparency and accountability 
in promoting open markets. The statistical summary and boxplot further confirm 
this relationship, showing that nations with higher integrity categories 
consistently achieve higher average economic freedom scores.

**Do countries with greater trade openness experience higher economic freedom?**

```{r, warning = F}
# Calculate the correlation between Trade Freedom and Economic Freedom
trade_freedom_correlation <- cor(economic_freedom_data$Trade.Freedom,
                                 economic_freedom_data$X2019.Score,
                                 use = "complete.obs")

# Display the correlation value
trade_freedom_correlation
```

```{r, warning = F}
# Scatter plot of Trade Freedom vs Economic Freedom
ggplot(economic_freedom_data, aes(x = Trade.Freedom, y = X2019.Score)) +
  geom_point(color = "brown", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Trade Freedom vs Economic Freedom Score (2019)",
    x = "Trade Freedom Score (0–100)",
    y = "Economic Freedom Score (0–100)",
    caption = "Source: Heritage Foundation, Economic Freedom Index 2019"
  ) +
  theme_minimal()

```

```{r, warning = F}
# Create Trade Freedom categories (quintiles)
economic_freedom_data <- economic_freedom_data %>%
  mutate(TradeFreedom_Category = cut(Trade.Freedom,
                                     breaks = quantile(Trade.Freedom, probs = seq(0, 1, 0.2), na.rm = TRUE),
                                     include.lowest = TRUE,
                                     labels = c("Low Trade Freedom",
                                                "Lower-Middle Trade Freedom",
                                                "Upper-Middle Trade Freedom",
                                                "High Trade Freedom",
                                                "Very High Trade Freedom")))

# Summary of Economic Freedom by Trade Freedom category
trade_freedom_summary <- economic_freedom_data %>%
  group_by(TradeFreedom_Category) %>%
  summarise(
    Mean_Freedom = mean(X2019.Score, na.rm = TRUE),
    SD_Freedom   = sd(X2019.Score, na.rm = TRUE),
    Min_Freedom  = min(X2019.Score, na.rm = TRUE),
    Max_Freedom  = max(X2019.Score, na.rm = TRUE)
  )

# Display the summary
trade_freedom_summary

```

```{r, warning = F}
# Boxplot of Economic Freedom by Trade Freedom category
ggplot(economic_freedom_data,
       aes(x = TradeFreedom_Category, y = X2019.Score, fill = TradeFreedom_Category)) +
  geom_boxplot() +
  labs(
    title = "Economic Freedom Score by Trade Freedom Category (2019)",
    x = "Trade Freedom Category",
    y = "Economic Freedom Score (0–100)",
    fill = "Trade Freedom Category"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

```

Conclusion

The results of the analysis show a strong positive correlation between trade 
freedom and economic freedom. The scatterplot demonstrates that countries with 
greater trade openness tend to achieve higher levels of economic freedom, 
highlighting the importance of free trade policies and reduced tariff barriers. 
The boxplot further supports this relationship, showing that nations in higher 
trade freedom categories consistently exhibit higher average economic freedom 
scores.

**Are there regional differences in economic freedom scores across the world?**

```{r, warning = F}
# Group data by region and summarize economic freedom scores
regional_freedom_summary <- economic_freedom_data %>%
  group_by(Region) %>%
  summarise(
    Mean_Freedom = mean(X2019.Score, na.rm = TRUE),
    SD_Freedom   = sd(X2019.Score, na.rm = TRUE),
    Min_Freedom  = min(X2019.Score, na.rm = TRUE),
    Max_Freedom  = max(X2019.Score, na.rm = TRUE)
  ) %>%
  arrange(desc(Mean_Freedom))

# Display the summary
regional_freedom_summary

```

```{r, warning = F}
# Boxplot of Economic Freedom Scores by Region
ggplot(economic_freedom_data,
       aes(x = reorder(Region, X2019.Score, FUN = median),
           y = X2019.Score,
           fill = Region)) +
  geom_boxplot() +
  labs(
    title = "Economic Freedom Scores by Region (2019)",
    x = "Region",
    y = "Economic Freedom Score (0–100)",
    fill = "Region"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Conclusion

Regional Variability: There are notable differences in economic freedom across 
regions. Europe has the highest mean economic freedom score, followed by the 
Middle East and North Africa, and the Asia-Pacific region. In contrast, 
Sub-Saharan Africa ranks the lowest, showing limited economic freedom overall.

Regions with High Economic Freedom: Europe stands out with the highest average 
score (68.6), reflecting strong institutions, open markets, and effective 
property rights protections. These regions typically benefit from transparent 
governance and stable regulatory environments that promote private enterprise.

Regions with Low Economic Freedom: Sub-Saharan Africa (mean 54.2) and the 
Americas (mean 59.6) have lower median economic freedom scores. This may be 
linked to greater government intervention, weaker property rights enforcement, 
and political instability that constrain market efficiency.

Variability within Regions: The Asia-Pacific and Americas regions exhibit wider 
ranges of scores, suggesting greater internal diversity. Some economies are 
highly open and free, while others face regulatory and institutional challenges 
that limit freedom.

**What patterns or similarities emerge between countries with high and low**
**economic freedom?**

```{r, warning = F}
# Select the top 5 countries with the highest economic freedom
top_5_freedom <- economic_freedom_data %>%
  arrange(desc(X2019.Score)) %>%
  select(Country, X2019.Score, Property.Rights, Government.Integrity, Tax.Burden,
         Business.Freedom, Trade.Freedom, Investment.Freedom, Financial.Freedom) %>%
  head(5)

# Select the bottom 5 countries with the lowest economic freedom
bottom_5_freedom <- economic_freedom_data %>%
  arrange(X2019.Score) %>%
  select(Country, X2019.Score, Property.Rights, Government.Integrity, Tax.Burden,
         Business.Freedom, Trade.Freedom, Investment.Freedom, Financial.Freedom) %>%
  head(5)

# Display both tables
top_5_freedom
bottom_5_freedom

```

```{r, warning = F}
# Calculate the mean and standard deviation for each factor for both groups
freedom_comparison <- data.frame(
  Factor = c("Property.Rights", "Government.Integrity", "Tax.Burden", "Business.Freedom",
             "Trade.Freedom", "Investment.Freedom", "Financial.Freedom"),
  High_Freedom_Mean = sapply(top_5_freedom[3:9], mean, na.rm = TRUE),
  High_Freedom_SD = sapply(top_5_freedom[3:9], sd, na.rm = TRUE),
  Low_Freedom_Mean = sapply(bottom_5_freedom[3:9], mean, na.rm = TRUE),
  Low_Freedom_SD = sapply(bottom_5_freedom[3:9], sd, na.rm = TRUE)
)

# Display the summary comparison
freedom_comparison

```

```{r, warning = F}
# Melt the data for easier plotting
freedom_comparison_melted <- melt(freedom_comparison, id.vars = "Factor",
                                  variable.name = "Group", value.name = "Value")

# Plot comparison of key factors: High vs Low Economic Freedom
ggplot(freedom_comparison_melted, aes(x = Factor, y = Value, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of Key Factors: High vs Low Economic Freedom Countries",
       x = "Factors",
       y = "Mean Score (0–100)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Conclusion

Economic Institutions: Countries with high economic freedom tend to have 
stronger institutions, reflected in significantly higher property rights and 
government integrity scores. These nations generally maintain transparent legal 
systems and low levels of corruption, creating a stable environment for economic activity.

Business and Trade Environment: High-freedom countries exhibit more open 
markets, with higher trade, investment, and financial freedom scores. These 
factors suggest that fewer restrictions on capital flows and international trade 
foster greater overall economic freedom.

Taxation and Regulation: Low-freedom countries typically have higher tax burdens 
and more restrictive regulatory frameworks, which limit private enterprise and discourage investment.

Overall Patterns: In short, the data indicate that economic openness, 
institutional integrity, and protection of property rights are key commonalities 
among high-freedom nations, while excessive government intervention and weak 
rule of law are consistent traits of low-freedom countries.

# Final Policy Implications

The results of this analysis highlight that higher economic freedom is most strongly associated with strong institutions—particularly well-protected property rights, judicial effectiveness, and government integrity. Regions with lower freedom often face structural challenges such as corruption, weak governance, and trade barriers. Policymakers should prioritize institutional reforms that strengthen rule of law, enhance transparency, and promote open markets. Encouraging stable regulatory frameworks and reducing trade restrictions can help countries move toward greater economic freedom and long-term prosperity.